
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% PREAMBLE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%% CLASS %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\documentclass[10pt,letterpaper]{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\usepackage[debugshow, final]{graphicx}
\usepackage{amsmath, amssymb, amsthm, amsfonts} 
\usepackage{enumerate}
\usepackage{fancybox} 
\usepackage{ltablex}
\usepackage{fancyhdr}
\usepackage{setspace} 
% \usepackage[top=1in,left=1in,bottom=1in,right=1in]{geometry} 
\usepackage{natbib}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{multicol}
\usepackage{textcomp} 
\usepackage{pdflscape}
\usepackage{lscape}
\usepackage[table]{xcolor}
\usepackage[T1]{fontenc}
\usepackage{subfigure}
\usepackage{hyperref}
\usepackage{cclicenses}
\usepackage[all]{hypcap}
\usepackage{listings}
% \usepackage{pxfonts}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%% SETTINGS %%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\pagestyle{fancy}
% \lhead{}
% \chead{}
% \rhead{}
% \cfoot{\thepage}

\author{Prepared for the starlab by J.P.\ Olmsted}

\title{\tsl{}'s Introduction to \R: A Short Course} 

\date{Updated: \today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%% CUSTOM MACROS %%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\let\margin\marginpar
\newcommand\myMargin[1]{
  \margin{\singlespace\raggedright\scriptsize
    \textsf{#1}}
} 
\renewcommand{\marginpar}[1]{\myMargin{#1}}

\newcommand\blurb[1]{
  \begin{center}
    \parbox[h]{.75\linewidth}{
      \footnotesize \singlespace
      #1  
    }
  \end{center}
}

\newcommand\R[0]{\texttt{R}}

\newcommand\tsl[0]{\textbf{the star lab}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% CONTENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\begin{document}

\begin{titlepage}
\maketitle 
\tableofcontents
\end{titlepage}

\section*{Introduction to this Short Course}

\paragraph{About.}This document was prepared by the author during his
tenure as \tsl{} Fellow for the incoming graduate student cohort in
the Political Science department at the University of
Rochester. Previous versions of the short course and the corresponding
materials are available at \tsl{}'s website
(\url{http://www.rochester.edu/college/psc/thestarlab/main/index.php}). This
document is an adaptation of Arthur Spirling's version. In addition to
the supplemental files used for the various exercises included in this
version, this document is available from both the starlab's website
and the author's website
(\url{http://www.rochester.edu/college/gradstudents/jolmsted/}).

\paragraph{Rights.}This document is released under the Creative
Commons Attribution license \by.

\paragraph{Comments.}If you have any comments, questions, or concerns
regarding this document please contact Jonathan Olmsted
(\url{jpolmsted@gmail.com}).

\paragraph{Description of Course.} This short course was covered in
five days. Parts 1 and 2 are covered in one day and the remaining
parts are covered on their own days. Part 1 discusses the whats and
whys of \R{}. Part 2 is a brief introduction to the fact that
everything in \R{} is an object and that objects have traits like
\textit{classes} and \textit{modes}. Part 3 is the first journey into
practice with the \R{} language. We walk through an example of how to
create and manipulate various objects in a way related to 

\pagebreak

\input{./1_Introduction/1_introduction.tex} 

\clearpage
\pagebreak

\input{./2_Fundamentals/2_fundamentals.tex}

\clearpage
\pagebreak

\input{./3_Case1/3_Case1.tex}

\clearpage
\pagebreak

\input{./4_Case2/4_Case2.tex}

\clearpage
\pagebreak

\input{./5_Programmation/5_programmation.tex}

\clearpage



%% \bibliographystyle{}
%% \bibliography{}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 


